---
title: "clean"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
library(tidyr)
library(dplyr)
library(ggplot2) 
library(ggmap)
library(geosphere)
library(psych)
register_google(key="AIzaSyDOHYjR93Vi0ols4DpE88pdPOppaO_aShg")

```

## CLEAN DATA

The python scripts read the data from [Google Big Query](https://bigquery.cloud.google.com) and creates four csv files

[yellow_2014.csv](https://github.com/celinakhalife/capstoneProject/blob/master/yellow_2014.csv)
[yellow_2015.csv](https://github.com/celinakhalife/capstoneProject/blob/master/yellow_2015.csv)
[green_2014.csv](https://github.com/celinakhalife/capstoneProject/blob/master/green_2014.csv)
[green_2015.csv](https://github.com/celinakhalife/capstoneProject/blob/master/green_2015.csv)

Since I'm dealing with a large dataset, dropping empty fields rows will not impact my modal. In fact, I will be able to create a more truthful analysis and prediction modal.
That's why, I decided to drop the records with missing fields like pickup_longitude, pickup_latitude, dropoff_longitude, dropoff_latitude, fare_amount, trip_distance and pickup_datetime.

As a first step, I decided to visualize the selected pickup_longitude and pickup_latitude using the `ggplot` library.

```{r}
yellowData2014 <- read.csv('../yellow_2014.csv', stringsAsFactors = FALSE)
plot <- ggplot(yellowData2014, aes(x=pickup_longitude, y=pickup_latitude)) +
  geom_point(size=0.06)
plot
```

As you can see, it appears that some locations are way off. To remove those records from the dataset, I calculated the distance as the Crow Flies of each pickup location to the center of New York City and removed all rows with a distance higher than 100miles.

```{r}
newYorkLatLng = c(-73.935242, 40.730610)
yellowData2014 <- yellowData2014  %>% mutate(
  distance_from_center = mapply(function(lg, lt) distm(newYorkLatLng, c(lg, lt), fun=distHaversine), pickup_longitude, pickup_latitude)/ 1609
)
yellowData2014 <- yellowData2014[yellowData2014$distance_from_center < 100,]
```

```{r, echo=FALSE}
yellowData2014$pickup_latitude <-  round(yellowData2014$pickup_latitude, 4)
yellowData2014$pickup_longitude <-  round(yellowData2014$pickup_longitude, 4)

min_lat <- min(yellowData2014$pickup_latitude)
max_lat <- max(yellowData2014$pickup_latitude)
min_long <- min(yellowData2014$pickup_longitude)
max_long <- max(yellowData2014$pickup_longitude)
```

```{r}
plot <- ggplot(yellowData2014, aes(x=pickup_longitude, y=pickup_latitude)) +
  geom_point(size=0.06) +
  scale_x_continuous(limits=c(min_long, max_long)) +
  scale_y_continuous(limits=c(min_lat, max_lat)) 

plot
```

Displaying the pickup longitude and latitude of yellow taxi 2014 on New York city using `ggmap`

```{r}

ggmap(get_map("New York",
              zoom = 12, scale = "auto",
              source = "google"),
      extent="device",
      legend="topright"
) + geom_point(aes(x=pickup_longitude, y=pickup_latitude), 
               data=yellowData2014, 
               col="red", alpha=0.2
)
```

Running the same as above on the yellow 2015 data will result in the following error
```
yellowData2015 <- yellowData2015  %>% mutate(
  distance_from_center = mapply(function(lg, lt) distm(newYorkLatLng, c(lg, lt), fun=distHaversine), pickup_longitude, pickup_latitude)/ 1609
)
Error in .pointsToMatrix(y) : longitude < -360
```

I had to remove all latitude and longitude values that are not within the maximum bounds
* Latitude: -85 to +85
* Longitude: -180 to +180

```{r, echo=FALSE}
yellowData2015 <- read.csv('../yellow_2015.csv', stringsAsFactors = FALSE)

min_lat <- min(yellowData2015$pickup_latitude)
max_lat <- max(yellowData2015$pickup_latitude)
min_long <- min(yellowData2015$pickup_longitude)
max_long <- max(yellowData2015$pickup_longitude)
```
```{r}
yellowData2015 <- yellowData2015[yellowData2015$pickup_latitude >= -85 & yellowData2015$pickup_latitude <= 85,]
yellowData2015 <- yellowData2015[yellowData2015$pickup_longitude >= -180 & yellowData2015$pickup_longitude <= 180,]
```

Running the same script again would display the following plot

```{r, echo=FALSE}

  yellowData2015 <- yellowData2015  %>% mutate(
    distance_from_center = mapply(function(lg, lt) distm(newYorkLatLng, c(lg, lt), fun=distHaversine), pickup_longitude, pickup_latitude)/ 1609
  )
  
  yellowData2015 <- yellowData2015[yellowData2015$distance_from_center < 100,]

  yellowData2015$pickup_latitude <-  round(yellowData2015$pickup_latitude, 4)
  yellowData2015$pickup_longitude <-  round(yellowData2015$pickup_longitude, 4)
  
  min_lat <- min(yellowData2015$pickup_latitude)
  max_lat <- max(yellowData2015$pickup_latitude)
  min_long <- min(yellowData2015$pickup_longitude)
  max_long <- max(yellowData2015$pickup_longitude)
  
  plot <- ggplot(yellowData2015, aes(x=pickup_longitude, y=pickup_latitude)) +
    geom_point(size=0.06) +
    scale_x_continuous(limits=c(min_long, max_long)) +
    scale_y_continuous(limits=c(min_lat, max_lat)) 
  
  plot
 
```

Displaying the pickup longitude and latitude of yellow taxi 2015 on New York city using `ggmap`

```{r, echo=FALSE}
  ggmap(get_map("New York",
                zoom = 12, scale = "auto",
                source = "google"),
        extent="device",
        legend="topright"
  ) + geom_point(aes(x=pickup_longitude, y=pickup_latitude), 
                 data=yellowData2015, 
                 col="red", alpha=0.2
  )
```

I repeat the same process on the green taxi data-sets

Green taxi 2014

```{r, echo=FALSE}
  greenData2014 <- read.csv('../green_2014.csv', stringsAsFactors = FALSE)
  min_lat <- min(greenData2014$pickup_latitude)
  max_lat <- max(greenData2014$pickup_latitude)
  min_long <- min(greenData2014$pickup_longitude)
  max_long <- max(greenData2014$pickup_longitude)
  
  greenData2014 <- greenData2014[greenData2014$pickup_latitude >= -85 & greenData2014$pickup_latitude <= 85,]
  greenData2014 <- greenData2014[greenData2014$pickup_longitude >= -180 & greenData2014$pickup_longitude <= 180,]
  
  #calculate distance to new york lat lng
  newYorkLatLng = c(-73.935242, 40.730610)
  greenData2014 <- greenData2014  %>% mutate(
    distance_from_center = mapply(function(lg, lt) distm(newYorkLatLng, c(lg, lt), fun=distHaversine), pickup_longitude, pickup_latitude)/ 1609
  )
  
  #remove all rows with a distance greater than 100 miles
  greenData2014 <- greenData2014[greenData2014$distance_from_center < 100,]
  
  #round longitude and latitude with a 4 precision
  greenData2014$pickup_latitude <-  round(greenData2014$pickup_latitude, 4)
  greenData2014$pickup_longitude <-  round(greenData2014$pickup_longitude, 4)
  
  # set ggplot limits
  min_lat <- min(greenData2014$pickup_latitude)
  max_lat <- max(greenData2014$pickup_latitude)
  min_long <- min(greenData2014$pickup_longitude)
  max_long <- max(greenData2014$pickup_longitude)
  
  plot <- ggplot(greenData2014, aes(x=pickup_longitude, y=pickup_latitude)) +
    geom_point(size=0.06) +
    scale_x_continuous(limits=c(min_long, max_long)) +
    scale_y_continuous(limits=c(min_lat, max_lat)) 
  
  plot
```

```{r}

  # draw pickup data on google maps
  ggmap(get_map("New York",
                zoom = 10, scale = "auto",
                source = "google"),
        extent="device",
        legend="topright"
  ) + geom_point(aes(x=pickup_longitude, y=pickup_latitude), 
                 data=greenData2014, 
                 col="red", alpha=0.2
  ) 
```

Green taxi 2015


```{r, echo=FALSE}
  greenData2015 <- read.csv('../green_2015.csv', stringsAsFactors = FALSE)
  min_lat <- min(greenData2015$pickup_latitude)
  max_lat <- max(greenData2015$pickup_latitude)
  min_long <- min(greenData2015$pickup_longitude)
  max_long <- max(greenData2015$pickup_longitude)

  greenData2015 <- greenData2015[greenData2015$pickup_latitude >= -85 & greenData2014$pickup_latitude <= 85,]
  greenData2015 <- greenData2015[greenData2015$pickup_longitude >= -180 & greenData2014$pickup_longitude <= 180,]
  
  #calculate distance to new york lat lng
  newYorkLatLng = c(-73.935242, 40.730610)
  greenData2015 <- greenData2015  %>% mutate(
    distance_from_center = mapply(function(lg, lt) distm(newYorkLatLng, c(lg, lt), fun=distHaversine), pickup_longitude, pickup_latitude)/ 1609
  )
  
  #remove all rows with a distance greater than 100 miles
  greenData2015 <- greenData2015[greenData2015$distance_from_center < 100,]
  
  #round longitude and latitude with a 4 precision
  greenData2015$pickup_latitude <-  round(greenData2015$pickup_latitude, 4)
  greenData2015$pickup_longitude <-  round(greenData2015$pickup_longitude, 4)
  
  # set ggplot limits
  min_lat <- min(greenData2015$pickup_latitude)
  max_lat <- max(greenData2015$pickup_latitude)
  min_long <- min(greenData2015$pickup_longitude)
  max_long <- max(greenData2015$pickup_longitude)
  
  plot <- ggplot(greenData2015, aes(x=pickup_longitude, y=pickup_latitude)) +
    geom_point(size=0.06) +
    scale_x_continuous(limits=c(min_long, max_long)) +
    scale_y_continuous(limits=c(min_lat, max_lat)) 
  
  plot
```
```{r}

  # draw pickup data on google maps
  ggmap(get_map("New York",
                zoom = 10, scale = "auto",
                source = "google"),
        extent="device",
        legend="topright"
  ) + geom_point(aes(x=pickup_longitude, y=pickup_latitude), 
                 data=greenData2015, 
                 col="red", alpha=0.2
  )
```

I run the same process on the dropoff latitude and longitude of all four data-sets and write the clean data-sets in four csv files

[clean_yellow_2014.csv](https://github.com/celinakhalife/capstoneProject/blob/master/clean_yellow_2014.csv)
[clean_yellow_2015.csv](https://github.com/celinakhalife/capstoneProject/blob/master/clean_yellow_2015.csv)
[clean_green_2014.csv](https://github.com/celinakhalife/capstoneProject/blob/master/clean_green_2014.csv)
[clean_green_2015.csv](https://github.com/celinakhalife/capstoneProject/blob/master/clean_green_2015.csv)


## WRANGLE DATA

Extract Date & Time Attributes from pickup_datetime and dropoff_datetime fields

```
  data <- data %>% mutate(
    pickup_year = mapply(function(datetime) as.POSIXlt(datetime)$year+1900 , pickup_datetime),
    pickup_month = mapply(function(datetime) as.POSIXlt(datetime)$mon+1 , pickup_datetime),
    pickup_day = mapply(function(datetime) as.POSIXlt(datetime)$mday , pickup_datetime),
    pickup_hour = mapply(function(datetime) as.POSIXlt(datetime)$hour , pickup_datetime),
    pickup_minute = mapply(function(datetime) as.POSIXlt(datetime)$min , pickup_datetime)
  ) 

  data <- data %>% mutate(
    dropoff_year = mapply(function(datetime) as.POSIXlt(datetime)$year+1900 , dropoff_datetime),
    dropoff_month = mapply(function(datetime) as.POSIXlt(datetime)$mon+1 , dropoff_datetime),
    dropoff_day = mapply(function(datetime) as.POSIXlt(datetime)$mday , dropoff_datetime),
    dropoff_hour = mapply(function(datetime) as.POSIXlt(datetime)$hour , dropoff_datetime),
    dropoff_minute = mapply(function(datetime) as.POSIXlt(datetime)$min , dropoff_datetime)
    
    )
```
