---
title: "clean"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
library(tidyr)
library(dplyr)
library(ggplot2) 
library(ggmap)
library(geosphere)
```

## CLEAN DATA

The python scripts read the data from [Google Big Query](https://bigquery.cloud.google.com) and creates four csv files
* [yellow_2014.csv](https://github.com/celinakhalife/capstoneProject/blob/master/yellow_2014.csv)
* [yellow_2015.csv](https://github.com/celinakhalife/capstoneProject/blob/master/yellow_2015.csv)
* [green_2014.csv](https://github.com/celinakhalife/capstoneProject/blob/master/green_2014.csv)
* [green_2015.csv](https://github.com/celinakhalife/capstoneProject/blob/master/green_2015.csv)

Since I'm dealing with a large dataset, dropping empty fields rows will not impact my modal. In fact, I will be able to create a more truthful analysis and prediction modal.
That's why, I decided to drop the records with missing fields like pickup_longitude, pickup_latitude, dropoff_longitude, dropoff_latitude, fare_amount, trip_distance and pickup_datetime.

As a first step, I decided to visualize the selected pickup_longitude and pickup_latitude using the `ggplot` library.

```{r}
yellowData2014 <- read.csv('../yellow_2014.csv', stringsAsFactors = FALSE)
plot <- ggplot(yellowData2014, aes(x=pickup_longitude, y=pickup_latitude)) +
  geom_point(size=0.06)
plot
```

As you can see, it appears that some locations are way off. To remove those records from the dataset, I calculated the distance as the Crow Flies of each pickup location to the center of New York City and removed all rows with a distance higher than 100miles.

```{r}
newYorkLatLng = c(-73.935242, 40.730610)
yellowData2014 <- yellowData2014  %>% mutate(
  distance_from_center = mapply(function(lg, lt) distm(newYorkLatLng, c(lg, lt), fun=distHaversine), pickup_longitude, pickup_latitude)/ 1609
)
yellowData2014 <- yellowData2014[yellowData2014$distance_from_center < 100,]
```

```{r, echo=FALSE}
yellowData2014$pickup_latitude <-  round(yellowData2014$pickup_latitude, 4)
yellowData2014$pickup_longitude <-  round(yellowData2014$pickup_longitude, 4)

min_lat <- min(yellowData2014$pickup_latitude)
max_lat <- max(yellowData2014$pickup_latitude)
min_long <- min(yellowData2014$pickup_longitude)
max_long <- max(yellowData2014$pickup_longitude)
```

```{r}
plot <- ggplot(yellowData2014, aes(x=pickup_longitude, y=pickup_latitude)) +
  geom_point(size=0.06) +
  scale_x_continuous(limits=c(min_long, max_long)) +
  scale_y_continuous(limits=c(min_lat, max_lat)) 

plot
```

Displaying the pickup longitude and latitude on New York city using `ggmap`

```{r}

register_google(key=[GoogleMapsKey])
ggmap(get_map("New York",
              zoom = 12, scale = "auto",
              source = "google"),
      extent="device",
      legend="topright"
) + geom_point(aes(x=pickup_longitude, y=pickup_latitude), 
               data=yellowData2014, 
               col="red", alpha=0.2
)
```