---
title: "clean"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggmap)
library(geosphere)
library(psych)
library(rgdal)
library(sf)
register_google(key=[key])

```

## EXPLORE DATA


### Working with Yellow and Green datasets

I wanted to visualize the density of the `trip_distance`, I used `quantile` to set the `trip_distance` threshold with a 90% probability.

```{r}

taxiData <-
  read.csv('../clean_yellow_green_data.csv', stringsAsFactors = FALSE)

threshold <- quantile(taxiData$trip_distance, probs=0.9)

ggplot(taxiData, aes(x=trip_distance, fill=taxi_color, alpha=I(0.4))) +
  xlim(min(taxiData$trip_distance),max(taxiData$trip_distance)) +
  geom_density() +
  geom_vline(xintercept=threshold) +
  scale_fill_manual(values=c("#006600", "#e5e500")) +
  labs(fill = 'Taxi Color', x = 'Trip Distance (in miles)', y = 'Density')
```

The trip distance threshold is `r threshold` miles
I decided to create a discrete modal with the new trip distance threshold

```{r}
distance_type_label <- c('TRUE' = paste('Trip Distance > ', threshold), 'FALSE' = paste('Trip Distance <= ', threshold))
ggplot(transform(taxiData,
                 long_distance = trip_distance > threshold), aes(x=trip_distance, fill=taxi_color, alpha=I(0.4))) +
  geom_density() +
  scale_fill_manual(values=c("#006600", "#e5e500")) +
  facet_grid (~long_distance,scales = "free", labeller=as_labeller(distance_type_label)) +
  labs(fill = 'Taxi Color', x = 'Trip Distance (in miles)', y = 'Density')
```

Running the same logic on `fare_amount` values

```{r}

taxiData <-
  read.csv('../clean_yellow_green_data.csv', stringsAsFactors = FALSE)

threshold <- quantile(taxiData$fare_amount, probs=0.9)

ggplot(taxiData, aes(x=fare_amount, fill=taxi_color, alpha=I(0.4))) +
  xlim(min(taxiData$fare_amount),max(taxiData$fare_amount)) +
  geom_density() +
  geom_vline(xintercept=threshold) +
  scale_fill_manual(values=c("#006600", "#e5e500")) +
  labs(fill = 'Taxi Color', x = 'Trip Distance (in miles)', y = 'Density')
```

`fare_amount` discrete modal

```{r}
amount_type_label <- c('TRUE' = paste('Fare Amount > ', threshold), 'FALSE' = paste('Fare Amount <= ', threshold))
ggplot(transform(taxiData,
                 long_distance = trip_distance > threshold), aes(x=trip_distance, fill=taxi_color, alpha=I(0.4))) +
  geom_density() +
  scale_fill_manual(values=c("#006600", "#e5e500")) +
  facet_grid (~long_distance,scales = "free", labeller=as_labeller(amount_type_label)) +
  labs(fill = 'Taxi Color', x = 'Fare Amount', y = 'Density')
```


# Locations

To classify the trip pickup locations, I used the neighborhood grid provided by (data.cityofnewyork.us)[https://data.cityofnewyork.us/City-Government/Neighborhood-Tabulation-Areas/cpf4-rkhq].

```{r}
utils::unzip('../neighborhoodTabulationAreas.zip')
shp <- readOGR(dsn = file.path('../geo_export_3b02421d-4a31-4bc5-ad76-6d5084b470f5.shp'), stringsAsFactors = F)

map <- ggmap(get_map("New York",
                     zoom = 10, scale = "auto",
                     source = "google"),
             extent="device",
             legend="topright"
)+ geom_point(aes(x=pickup_longitude, y=pickup_latitude, color=taxi_color),
              data=taxiData,
               alpha=I(0.2)
)  +
  geom_polygon(data = shp, aes(x = long, y = lat, group = group), color = "black", fill = NA)
map + theme_void()
```

I projected the pickup location over the neighborhoods shapes and retrieved the neighborhood name and code. I created the same previous plot with neighborhood color indexes

```{r}
spacialTaxiData <- taxiData
coordinates(spacialTaxiData)=~pickup_longitude+pickup_latitude
nhood2 = spTransform(shp, CRS("+proj=longlat +datum=WGS84"))

proj4string(spacialTaxiData) = proj4string(nhood2)
spacialTaxiData$nhoodID <- over(spacialTaxiData, nhood2)
taxiData$ntaname <- unlist(spacialTaxiData$nhoodID['ntaname'])
taxiData$ntacode <- unlist(spacialTaxiData$nhoodID['ntacode'])

ggplot(taxiData, aes(x=pickup_longitude, y=pickup_latitude)) +
  geom_point(aes(color=ntacode, alpha=I(0.2)))+ theme(legend.position = "none")  +
  geom_polygon(data = shp, aes(x = long, y = lat, group = group), color = "black", fill = NA) +
  labs(x = 'Longitude', y = 'Latitude')
```

Summary extracted from the neighborhood data

Highest 15 neighborhoods

```{r}

nhoodSummary <- taxiData %>% group_by(ntacode) %>%
  summarise(count=n())
colnames(nhoodSummary) <- c("ntacode", "value")

first15NH <- nhoodSummary %>% arrange(desc(value)) %>% head(15)
d <- taxiData[taxiData$ntacode %in% first15NH$ntacode,]

ggplot(d, aes(x=ntacode, fill=taxi_color, alpha=I(0.4))) +
  geom_bar() +
  labs(x = 'Neighborhood Code', y = 'Trips Density', fill = 'Taxi Color') +
  scale_fill_manual(values=c("#006600", "#e5e500"))

```

Lowest 15 neighberhoods

```{r}

last15NH <- nhoodSummary %>% arrange(value) %>% head(15)
d <- taxiData[taxiData$ntacode %in% last15NH$ntacode,]

ggplot(d, aes(x=ntacode, fill=taxi_color, alpha=I(0.4))) +
  geom_bar() +
  labs(x = 'Neighberhood Code', y = 'Trips Density', fill = 'Taxi Color') +
  scale_fill_manual(values=c("#006600", "#e5e500"))

```

